---
title: "final_work_federico_loguercio"
author: "Federico Loguercio"
date: "12/21/2018"
output: html_document
---

```{r setup, echo = F}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = F, message = F)
options(repos=c(CRAN="https://cran.rediris.es/"))
```

```{r Instructions, echo=FALSE}

#R Markdown

#Download the nasdaq_symbols.xlsx file from the datasets area in the campus. 
#It contains symbols and companies names for the Nasdaq stock market.
#Create a report where you analize the relationship between the average daily 
#return and the volatility (sd), for each symbol.
#Is (daily) profitability related with (daily) volatility? 
#Create a clustering of companies, using the mean and sd of daily returns as the spliting variables. 
#Analize the results.
#Important: Due date is friday 21st dec. at 11.59 PM. Submit a single file (or zipped folder) with the naming structure: final_work_firstName_lastName

```


```{r packages, echo=FALSE}
rm(list=ls())
# Set wd to folder containing nasdaq_symbols.xlsx
wd <- getwd()

#Install necessary packages if not downloaded:
if(!"quantmod" %in% installed.packages()) {
  install.packages("quandmod")
} 
if(!"readxl" %in% installed.packages()) {
  install.packages("readxl")}
if(!"data.table" %in% installed.packages()) {
  install.packages("data.table")}
if(!"ggplot2" %in% installed.packages()) {
  install.packages("ggplot2")}
if(!"plotly" %in% installed.packages()) {
  install.packages("plotly")}
if(!"wesanderson" %in% installed.packages()) {
  install.packages("wesanderson")}
```

### Is daily profitability related to volatility?
Read data for a set of nasdaq stocks, since 2010:
```{r readdata}
library(data.table)
library(readxl)
library(quantmod)
nasdaq_symbols <- as.data.table(read_xlsx(paste0(wd,"/nasdaq_symbols.xlsx")))

# Get stock data for all tickers, lapplying getSymbols to every ticker. Auto.assign in order to return results
stocks <- lapply(nasdaq_symbols$Symbol, function(x) getSymbols(x, auto.assign=F, from = "2010/01/01"))
```


Prepare data into the desired format, with two values for each stock, average daily return and volatility:
```{r preparedata}
# Extract closing price, keep xts format:
# Inside: lApply quantmod function Cl to extract the closing price from every list in stocks
# Outside: Apply merge to the resulting list (like lapply, but taking instead of returning a list)
closing_price <- do.call(merge, lapply(stocks, function(x) Cl(x)))

# Create object with daily returns
# Inside: lApply quantmod's dailyReturn function to each element in closing_price
# Outside: Apply xts' merge.xts function (package loaded with quantmod) to every object in list
# just created, returning another xts
daily_returns <- do.call(merge.xts, lapply(closing_price, dailyReturn))

# Assign original column names to daily_returns. Column order was not changed
names(daily_returns) <- names(closing_price)

# Calculate average daily return and volatilty (as standard deviation)
# for each stock
mean_sd <- as.data.table(sapply(daily_returns, function(x) c("avg_daily_return"= mean(x,na.rm=T), "volatility" = sd(x, na.rm=T))))

# Transpose data in order to have a row for mean and one for sd using
#data.table function transpose, assigning colnames:
final <- transpose(mean_sd, fill=NA, ignore.empty=FALSE)
colnames(final) <- c('avg_daily_return', 'volatility')
# Create variable with ticker, cropping only the ticket (string before the dot)
final$ticker <- names(mean_sd)
final$ticker <- sub("\\..*", "", final$ticker)

# Scale each numeric variable for clustering with mean=0, sd=1
data_cluster<-data.table(scale(final[, final[,c('avg_daily_return', 'volatility')]]))
```


Is there a correlation between daily profitability and volatility?
```{r correlation, echo = F}
# Call necessary libraries
library(ggplot2)
library(plotly)
# Are volatility and return correlated?
print(paste("Pearson correlation coefficient =", cor(final$volatility, final$avg_daily_return)))

# Plot Return on volatility, adding a fitted linear model with shaded confidence intervals
ggplotly(ggplot(final, aes(volatility, avg_daily_return, text = ticker)) +
  geom_point(color="darkred") +
  labs(title = 'Return and Volatility', y = 'Avg. Daily Return', x = 'Volatility') +
  geom_smooth(data = final, method = lm, inherit.aes = F, aes(volatility, avg_daily_return),color="darkgreen"))

```
The graph suggests a positive relationship between volatility and average daily returns. The simple linear model shown in green confirms this.


#### Are there different groups of stocks in terms of this relationship? A clustering analysis could yield some insights.
What is the ideal number of clusters for K-means clustering?
```{r numofclusters}
# Determine optimum numbers of clusters using ellbow method, plotting
# within cluster sum of square
set.seed(1337)
# Compute within cluster sum of square for k = 2 to k = 10
# nstart = 10: Try different initial centroids and choose best
# Apply the function kmeans using k ranging from 2 to 10, extract only
# the output element indicating total within sum of squares and assign to wss
wss <- sapply(1:10, function(k){kmeans(data_cluster, k, nstart=10)$tot.withinss})

#Plot the different wss
plot(1:10, wss,type="b", pch = 19, frame = FALSE, xlab="Number of clusters K", ylab="Total within-clusters sum of squares")
```

The graph suggests that the decrease in within cluster sum of squares is much lower with every additional cluster after k=2.


```{r clustering_2}
# Fit kmeans with 2 clusters
kmeans_2 <- kmeans(data_cluster, 2, nstart=10)

final$cluster_2 <- as.factor(kmeans_2$cluster)

# Load some nice colour palettes
library(wesanderson)

ggplotly(ggplot(final, aes(volatility, avg_daily_return, colour = cluster_2, text = ticker)) +
  geom_point() +
  labs(title = 'Return and Volatility, 2 Clusters', y = 'Avg. Daily Return', x = 'Volatility') +
  scale_color_manual(values=wes_palette(n=2, name="Darjeeling1"))+
  theme(legend.position = "none"))
```


Explore different numbers of clusters in order to analyse, if a number different from 2 yields clusters with clearer profiles
```{r clustering_3}
# Fit kmeans with 3 clusters
kmeans_3 <- kmeans(data_cluster, 3, nstart=10)

final$cluster_3 <- as.factor(kmeans_3$cluster)

ggplotly(ggplot(final, aes(volatility, avg_daily_return, colour = cluster_3, text = ticker)) +
  geom_point() +
  labs(title = 'Return and Volatility, 3 Clusters', y = 'Avg. Daily Return', x = 'Volatility') +
  scale_color_manual(values=wes_palette(n=3, name="Darjeeling1"))+
  theme(legend.position = "none"))
```

```{r clustering_4}
# Fit kmeans with 3 clusters
set.seed(1337)
kmeans_4 <- kmeans(data_cluster, 4, nstart=10)

final$cluster_4 <- as.factor(kmeans_4$cluster)

ggplotly(ggplot(final, aes(volatility, avg_daily_return, colour = cluster_4, text = ticker)) +
  geom_point() +
  labs(title = 'Return and Volatility, 4 Clusters', y = 'Avg. Daily Return', x = 'Volatility') +
  scale_color_manual(values=wes_palette(n=4, name="Darjeeling1"))+
  theme(legend.position = "none"))

# Create file for shiny app
write.csv(final, "federico_data.csv")
```

In general, it can be observed that K-means splits the stocks into low volatility and low returns vs high volatility and high returns, with various levels according to the number of clusters. When adding a third cluster, the stocks with negative average daily returns are grouped into one separate cluster, which could be interpreted as an outliers-cluster. Nothing unexpected is revealed.
From the four-cluster solution the following could be inferred: The red cluster is made up of negative averagereturn stocks which should be avoided. The orange stocks are a good low-risk option: returns are decent, with low risk (volatility). The yellow cluster is a more high-yield set of stocks, with high returns at the cost of high volatility. The green cluster is somewhat in between.
